### 2025-10-25 thesis/implementation/index.tex
More impl stuff

The \texttt{selection} algorithm implementation follows a two-stage approach,
split into initial candidate retrieval through vector search and subsequent
reweighting using the above described weights and a precomputed WWL distance
index from \texttt{natural-graphs}.

>>>>

The \texttt{selection} algorithm implementation follows a two-stage approach,
split into initial candidate retrieval through vector search (using the
\texttt{Vector} struct and SQLite) and subsequent reweighting using the above
described weights and a precomputed WWL distance index from
\texttt{natural-graphs}.

\subsubsection{Schema Subsetting System ($\phi$)}

% - Execution-based validation vs heuristic approaches
% - Dynamic pruning: 500-800ms for 20+ table schemas vs 10-15ms heuristic
% - In-memory database creation and testing for candidate queries
% - Schema dependency analysis challenges (nested subqueries, CTEs)
% - Correctness vs performance trade-off: eliminates pipeline failures

The schema subsetting system is implemented in
\texttt{natural-inference::pipeline::subsetting} as
\texttt{SchemaSubsetter} struct. The core algorithm is the \texttt{optimize}
method (lines 21-47) which performs execution-based validation to determine
which tables are crucial for the query candidates provided.

\paragraph{Execution-based Validation}

The implementation uses a trial-and-error approach where for each query
candidate and each table in the schema, it creates an in-memory SQLite database
using \texttt{Connection::open\_in\_memory()} and populates it's schema with all
tables except the current one being tested. It then attempts to prepare every
query candidate against this reduced schema.

If the query preparation fails, this indicates the removed table is crucial
for the query, and it gets added to the list of crucial tables.
This process is repeated for all query-table combinations to build a minimal
schema containing only essential tables for the provided set of query candidates.

\paragraph{Performance and Trade-off Characteristics}

The algorithm creates a new in-memory database for each table-query
combination, leading to $O(queries \times tables)$ complexity.

\textcolor{red}{For schemas with XXX+ tables and multiple query candidates, this results in XXX-XXXms processing time
compared to 10-15ms for heuristic approaches.}

However, this execution-based approach provides correctness guarantees by
eliminating pipeline failures caused by missing table dependencies. The
method handles complex SQL dependencies including nested subqueries, CTEs,
and foreign key relationships that heuristic approaches often miss.

\paragraph{Pipeline Integration}

The \texttt{SchemaSubsetter} is used prior to prompting the model using
in-context learning to ensure that the model context is used optimally.
The \texttt{optimize} method returns a \texttt{Schema} object
containing only the tables identified as crucial through execution testing
which is in turn processed by the subsequent pipeline stages like generation
and refinement.
